SSA based global code motion of identical computations

### 2017-04-17 acmart-pacmpl-template.tex
mover partition to profitability

\subsection{Finding candidates to move}
\label{subsec:finding-candidates}
The first step is to find a set of instructions that perform identical
computations: this is performed by a linear scan of all instructions of the
program and classifying all instructions by their value given by \GVN{}. We
could compute available and anticipable sets as computed by GCC's code-hoisting
but that would be a lot of data structures to maintain at each basic block
level. For GCC it makes sense because their code-hoisting is integrated with
GVN-PRE which already has those data structures available.

@@

Once the instructions have been classified into congruence classes
\cite{briggs1997} i.e., a set of instructions which compute the same value, we
compute for each group of congruent instructions, a point in the program that is
both legal and profitable for the instructions to be moved to.

@@

\subsubsection{Legality of hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the addresss if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case we discard the
candidate. We use the MemorySSA infrastructure of \LLVM{} to iterate on the
use-def chains available for memory references.

\subsubsection{Legality of hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads using the MemorySSA of \LLVM{}. We check that the operands of the store
instruction are available at the hoisting point, that there are no aliasing
loads or store along the path from the current position to the hoisting point.

@@

\subsubsection{Legality of sinking expressions}
One might wonder that if a set of expressions compute the same value then why
not just hoist them. The problem is, sometimes, hoisting is not upward-safe
\ref{click1995global} e.g., if the expressions are in a landing pad
etc. Sometimes sinking may reduce the register pressure e.g., when the use
operands are not kills.

%TODO: Explain why all the identical computation cannot be hoisted.
For sinking, higher ranked expressions would be sunk first. And it would be
illegal to sink higher ranked identical expressions if they are not anticipable
in the common post-dominator. For example:

@@

In this example (c1, c2) or (d1, d2) could be sunk first. Since (c1, c2) depend
on i1 and i2 respectively which are also in their original basic blocks, c1 and
c2 are not anticipable in B3. So without knowing the sinkability of `i1' and
`i2' it would be illegal to sink (c1, c2) to B3. On the other hand (d1, d2) can
safely be sunk because their operands are readily available at the sink point
i.e., B3. It should also be noted that, just because the expressions are
identical and operands are available, it still requires a unique post-dominating
PHI to use the exact same values to be legally sinkable.

A general global scheduling algorithm also requires checks for undefinedness
when introducing a new computation along a path. Since only very busy
expressions are moved in the current implementation, there is no need to check
for undefinedness resulting due to movement of instruction. This simplifies the
implementation.

\subsubsection{Partitioning the list of hoistable candidates to maximize hoisting}
\label{subsec:partition}
%The partition algorithm is described as follows:
In the approach, described in Section~\ref{subsec:optimistic}, it is possible
that a common hoisting point of all the instructions is either too far away, or
not legally possible. In these cases, it is still possible to `partially' hoist
a subset of instructions by splitting the set of candidates and finding a closer
hoisting point for each subset.

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and
for some reason they cannot be hoisted together at B0, then we partition the set
of hoistable candidates in their DFSIn order. In this case the DFSIn ordering
would be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be
hoisted at B1 and those in B5, B6 to be hoisted at B2. Even if (B3, B4) could
not be hoisted, for some reason, to B1, (B5, B6) can still be hoisted -- if
legal and profitable -- to B2.

@@

Hoisting loads/stores across calls also require precise analysis of all the
memory addresses accessed by the call. The current implementation being an
intraprocedural pass, cannot hoist aggressively across calls. In the presence of
pure calls, loads can be hoisted but stores can't. Also, if the call throws
exceptions, or if it may not return, nothing can be hoisted.

>>>>

\subsection{Finding candidates to move}
\label{subsec:finding-candidates}
The first step is to find a set of congruent instructions
\cite{briggs1997}. This is performed by a linear scan of all instructions of the
program and classifying them by their value numbers. We could compute available
and anticipable sets as computed by GCC's code-hoisting but that would be a lot
of data structures to maintain at each basic block level. For GCC it makes sense
because their code-hoisting is integrated with GVN-PRE which already has those
data structures available.

@@

Once the instructions have been classified into congruence classes, we compute
for each group of congruent instructions, a point in the program that is both
legal and profitable for the instructions to be moved to.

@@

\subsubsection{Legality of hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the addresss if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case the candidate is
discarded. To iterate on the use-def chains for memory references the MemorySSA
infrastructure of \LLVM{} is used.

\subsubsection{Legality of hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads using the MemorySSA of \LLVM{}. We check that the operands of the store
instruction are available at the hoisting point, and that there are no aliasing
loads or store along the path from the current position to the hoisting point.

@@

Hoisting loads/stores across calls also require precise analysis of all the
memory addresses accessed by the call. The current implementation being an
intraprocedural pass, cannot hoist aggressively across calls. In the presence of
pure calls, loads can be hoisted but stores can't. Also, if a call throws
exceptions, or if it may not return, nothing can be hoisted across that call.

\subsubsection{Legality of sinking expressions}
Sometimes, hoisting is not upward-safe \ref{click1995global} e.g., if the
expressions are in a landing pad etc., in that case sinking of those expressions
may reduce code size. Sinking may also reduce the register pressure in some
cases e.g., when the use operands are not kills.For sinking, higher ranked
expressions would be sunk first. And it would be illegal to sink higher ranked
identical expressions if they are not anticipable in the common
post-dominator. For example:

@@

In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
(c1, c2) depend on i1 and i2 respectively which are also in their original basic
blocks, c1 and c2 are not anticipable in B3. So without knowing the sinkability
of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
(d1, d2) can safely be sunk because their operands are readily available at the
sink point i.e., B3. It should also be noted that, just because the expressions
are identical and operands are available, it still requires a unique
post-dominating PHI to use the exact same values to be legally sinkable.

A general global scheduling algorithm also requires checks for undefinedness
when introducing a new computation along a path. Since only very busy
expressions are moved in the current implementation, there is no need to check
for undefinedness resulting due to movement of instructions. This simplifies the
implementation.

@@

\subsubsection{Partitioning the list of hoistable candidates to maximize hoisting}
\label{subsec:partition}
%The partition algorithm is described as follows:
In the approach, described in Section~\ref{subsec:optimistic}, it is possible
that a common hoisting point for all the instructions does not satisfy legality
or profitability checks. In these cases, it is still possible to `partially'
hoist a subset of instructions by splitting the set of candidates and finding a
closer hoisting point for each subset.

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
for some reason, they cannot be hoisted at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be hoisted at
B1 and those in B5, B6 to be hoisted at B2. Even if (B3, B4) could not be
hoisted, for some reason, to B1, (B5, B6) can still be hoisted -- if legal and
profitable -- to B2.
